Publishing Datasets about Building Designs \\Based on a Multilayer Building Ontology

### 2015-04-29 article-02-main.tex
Wrote about defined, entity types. Removed listing captions to save space.

@@

\begin{ontologyRule}[OWL profile, ifcOWL layers: all]
All ifc\-OWL ontologies must be based on OWL 2 RL profile\footnote{Our IFC-EXPRESS-to-OWL converter tool which is available on http://drumbeat.cs.hut.fi also allows choosing other OWL profiles: OWL Lite, OWL DL, OWL Full, OWL 2 EL, OWL 2 QL, OWL 2 Full}.
\end{ontologyRule}

This profile has been chosen because in contrast to OWL 2 EL and OWL 2 QL, it supports such needed things as: enumerations of individuals (see \ref{ifcOWL:enum-types}), named individuals (see \ref{ifcRDF:naming-individuals}), etc.\cite{w3c:owl2-profiles}.

\begin{ontologyRule}[Domain and range information, ifcOWL layers: all]
Property domain and range constraints specified by \name{rdfs:domain} and \name{rdfs:range} should not be included in ifcOWL ontologies.
\end{ontologyRule}

@@

\begin{ontologyRule}[Namespaces, ifcOWL layers: all]
All constructs related purely to EXPRESS, STEP or IFC specifications must be belong to namespaces with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly. At this moment the namespaces are defined as in Listing \ref{lst:ifcOWL-namespaces}. Instead of \name{IFCXXX} there should be IFC versions, e.g. \name{IFC2x3} or \name{IFC4}.
\end{ontologyRule}

\begin{lstlisting}[caption={Namespace definitions},label=lst:ifcOWL-namespaces]
@prefix expr: <http://drumbeat.cs.hut.fi/owl/EXPRESS#>
@prefix step: <http://drumbeat.cs.hut.fi/owl/STEP#>
@prefix ifc: <http://drumbeat.cs.hut.fi/owl/IFCXXX#>
\end{lstlisting}

@@

@@

Simple data\-types built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. However, type \name{LOGICAL} which has three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} is considered by us as an enumeration data\-type (see \ref{ifcOWL:enum-types}). The same to type \name{BOOLEAN} which is like a primitive boolean data\-type but semantically close and met with the same frequency in IFC4 as \name{LOGICAL}.

On the other hand, there are two defined data\-types in IFC (not in EXPRESS) that also can be referred to the simple-data-type group: \name{Ifc\-Time\-Stamp} which is a so-called TimeSpan data\-type and can be easily converted to \name{xsd:date\-Time}, and \name{Ifc\-Globally\-Unique\-Id} which represents set of compressed Globally Unique Identifiers (GUIDs), i.e. strings with fixed length (see Listing \ref{lst:ifcOWL-additional-simple-types}).

\begin{lstlisting}[caption={Two defined data\-types considered as simple data\-types}, label=lst:ifcOWL-additional-simple-types]
(* number of seconds elapsed since the beginning of the year 1970 *)
TYPE IfcTimeStamp = INTEGER; END_TYPE;

(* compressed GUID *)
TYPE IfcGloballyUniqueId = STRING(22) FIXED; END_TYPE;
\end{lstlisting}

\begin{ontologyRule}[Simple data\-types, ifcOWL layers: all]
All simple data\-types except \emph{BOOLEAN} and \emph{LOGICAL} are the same as their most similar primitive XSD data\-types which are supported by OWL 2 \cite{w3c:owl2-profiles} and are among preferred standard data\-types in Linked Data Platform \cite{w3c:ldp-best-practices} (see Listing \ref{lst:ifcOWL-simple-types}). However, it is recommended to use directly XSD s like (e.g. \name{xsd:string}) instead of the EXPRESS data\-type constructs (e.g. \name{expr:STRING}).
\end{ontologyRule}

% For \name{STRING} and \name{Ifc\-Globally\-UniqueId} it is \name{xsd:\-string}, for \name{BINARY} -- \name{xsd:\-hexBinary}, for \name{INTEGER} -- \name{xsd:\-integer}, for \name{REAL} and \name{NUMBER} -- \name{xsd:\-double}, for \name{Ifc\-Time\-Stamp} -- \name{xsd:\-dateTime}.

\begin{lstlisting}[caption={Simple data types}, label=lst:ifcOWL-simple-types]
expr:STRING, ifc:IfcGlobalUniqueId owl:sameAs xsd:string .
expr:BINARY owl:sameAs xsd:hexBinary .
expr:INTEGER owl:sameAs xsd:integer .
expr:REAL, expr:NUMBER owl:sameAs xsd:double .
ifc:IfcTimeStamp owl:sameAs xsd:dateTime .
\end{lstlisting}

`Most similar XSD data\-types' mean the types from XSD schema that conform to the value spaces of the original EXPRESS and IFC data\-types better than other ones. For instance, type \name{xsd:double} is more similar to \name{REAL} than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data\-type and supports scientific notation \cite{w3c:xsd,datapic:xsd}.

\begin{ontologyRule}[Simple data\-types with sizes, ifcOWL layers: all]If \name{STRING} and \name{BINARY} data\-types are specified with a size parameter, i.e. \name{STRING (max\-Length\-In\-Bytes)} or \name{BINARY(size\-In\-Bits)}, then they will be named in format \name{\textless type\_name \textgreater\textless size\textgreater} (e.g. \name{STRING22}). In layers \lite{} and \standard{} they are equivalent to data\-types without size. But in layer \advanced{}, in order to specify the size parameter, additional constraints must be added by using property \name{owl:onDatatype} and data\-type facets \name{xsd:maxLength} or \name{xsd:length} \cite{w3c:owl2-syntax,w3c:xmlschema11-2} (see example in Listing \ref{lst:ifcOWL-simple-types-with-size}).
\end{ontologyRule}

\begin{lstlisting}[caption={Size constraint of simple s},
label=lst:ifcOWL-simple-types-with-size]
# only for Lite, Standard layers
expr:STRING22 owl:sameAs expr:STRING .

# only for Advanced layer
expr:STRING22 owl:equivalentClass
    [   rdf:type        rdfs:Datatype ;
        owl:onDatatype  xsd:string ;
        owl:withRestrictions ( [ xsd:length "22"^^xsd:integer ] ) ] .
\end{lstlisting}

% \begin{lstlisting}[caption={},label=1]
% expr:STRING rdf:type owl:Class;
%     rdfs:subClassOf [
%         owl:Restriction;
%         owl:onProperty expr:hasValue;
%         owl:allValuesFrom xsd:string;
%     ].

% \begin{lstlisting}[caption={Simple data\-types as instances of \name{owl:Class}}, label=lst:ifcOWL-simple-types]
% expr:STRING a owl:Class;
%     rdfs:subClassOf [ a owl:Restriction;
%                       owl:onProperty rdf:value;
%                       owl:allValuesFrom xsd:double; ] .
% ifc:IfcTimeStamp a owl:Class;
%     rdfs:subClassOf [ a owl:Restriction;
%                       owl:onProperty rdf:value;
%                       owl:allValuesFrom xsd:dateTime; ]
% \end{lstlisting}

% The reason of using class boxing technique is to make simple data\-types be instances of \name{owl:Class}. Instead of that they could be either simply replaced with primitive XSD data\-types, or defined as shown in the Listing , but in these cases they will be instances of \name{rdfs:Datatype}.  and cannot be used for union classes

% \begin{lstlisting}[caption={Simple data\-types as instances of \name{rdfs:Datattype}},label=lst:ifcOWL-simple-types-2]
% # if LEVEL_OF_DETAIL is higher than STANDARD
% expr:STRING owl:equivalentClass [
%     rdf:type        rdfs:Datatype ;
%     owl:onDatatype  xsd:string ;
%     owl:withRestrictions ()
% ]
% \end{lstlisting}

% Other considered conversion options were:
% \begin{itemize}
% \item To replace simple data\-types with the primitive XSD data\-types. But in this case all select data\-types (see \ref{subsec:ifcOWL:select-types}) cannot be defined by means of property \name{owl:unionOf} the range of which

% all properties in any OWL profile except OWL Full must be either \name{owl:}
% \end{itemize}

% ifc:ifcLabel rdf:type owl:Class;
%     rdfs:subClassOf expr:STRING.
% \end{lstlisting}

% \begin{lstlisting}[caption={},label=1]

% % http://answers.semanticweb.com/questions/229/how-to-define-a-own-notation-data\-type

% http://semantic-web-grundlagen.de/wiki/Guide_to_OWL_2_for_OWL_1_users#Datatypes

% http://www2.informatik.hu-berlin.de/~wandelt/SW201213/9OWL2.pdf

% Using that in OWL 2 DL is no problem. If you want to stay within one of the profiles though note that defining data\-types through such restrictions isn't allowed in any of them. So for increased interoperability (also with OWL 1 and RDFS) you could just use xsd:string instead. That should be perfectly fine with SKOS.

% expr:STRING rdf:type rdfs:Datatype;
%  owl:onDatatype xsd:string;
%  owl:withRestrictions ().

% ifc:ifcLabel rdf:type owl:Class;
%     rdfs:subClassOf expr:STRING.
% \end{lstlisting}

@@

% \item \textbf{Defined or declared data type}: This further specializes other data\-types — e.g., define a datatype \emph{PositiveInteger} that is of type integer with a positive value.

\subsection{Entity Datatype Declarations}

\subsection{Enumeration Type Declarations}
\label{ifcOWL:enum-types}

>>>>

\begin{ontologyRule}[OWL profile]
\ifclite{} layer must be compatible with any OWL 2 DL profile: OWL 2 EL, OWL 2 QL or OWL 2 RL. \ifcstandard{}, \ifcadvanced{} must be compatible with OWL 2 RL\footnote{Our WebIFC converter tool allows forced choice of other OWL profiles for any layer: OWL Lite, OWL DL, OWL Full, and OWL 2 Full}.
\end{ontologyRule}

@@

OWL 2 RL has been chosen for \ifcstandard{}, \ifcadvanced{} because in contrast to OWL 2 EL and OWL 2 QL, it supports such needed things as: union data\-types (see \ref{subsec:ifcOWL-select-types}), enumerations of individuals (see \ref{subsec:ifcOWL-enum-types}), anonymous individuals (see \ref{subsec:ifcRDF-naming-individuals}), etc.\cite{w3c:owl2-profiles}.

\begin{ontologyRule}[Domain and range information]
Property domain and range constraints specified by \name{rdfs:domain} and \name{rdfs:range} should not be included in ifcOWL ontologies.
\end{ontologyRule}

@@

\begin{ontologyRule}[Namespaces]
All constructs related to EXPRESS, STEP or IFC specifications must be belong to namespaces with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly. At this moment the namespaces are defined as below (istead of \name{IFCXXX} there should be IFC versions, e.g. \name{IFC2x3} or \name{IFC4}):

%\begin{lstlisting}[caption={Namespace definitions},label=lst:ifcOWL-namespaces]
\begin{lstlisting}
@prefix expr: <http://drumbeat.cs.hut.fi/owl/EXPRESS#>
@prefix step: <http://drumbeat.cs.hut.fi/owl/STEP#>
@prefix ifc: <http://drumbeat.cs.hut.fi/owl/IFCXXX#>
\end{lstlisting}
\end{ontologyRule}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simple Datatype Declarations
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Simple data\-types built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. However, type \name{LOGICAL} which has three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} is considered by us as an enumeration data\-type (see \ref{subsec:ifcOWL-enum-types}). The same to type \name{BOOLEAN} which is like a primitive boolean data\-type but semantically close and met with the same frequency in IFC4 as \name{LOGICAL}.

\begin{ontologyRule}[Simple data\-types]
All simple data\-types except \emph{BOOLEAN} and \emph{LOGICAL} are declared as the same as their most similar primitive XSD data\-types which are supported by OWL 2 \cite{w3c:owl2-profiles} and are among preferred standard data\-types in Linked Data Platform \cite{w3c:ldp-best-practices}:

% \begin{lstlisting}[caption={Simple data types}, label=lst:ifcOWL-simple-types]
\begin{lstlisting}
expr:STRING, ifc:IfcGlobalUniqueId owl:sameAs xsd:string .
expr:BINARY owl:sameAs xsd:hexBinary .
expr:INTEGER owl:sameAs xsd:integer .
expr:REAL, expr:NUMBER owl:sameAs xsd:double .
\end{lstlisting}
\end{ontologyRule}

`Most similar XSD data\-types' mean the types from XSD schema that conform to the value spaces of the original EXPRESS and IFC data\-types better than other ones. For instance, type \name{xsd:double} is more similar to \name{REAL} than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data\-type and supports scientific notation \cite{w3c:xsd,datapic:xsd}. It is recommended to use XSD data\-type names directly when they are needed.

\begin{ontologyRule}[Simple data\-types with sizes, ifcOWL layers: all]If \name{STRING} and \name{BINARY} data\-types are specified with a size parameter, i.e. \name{STRING (max\-Length\-In\-Bytes)} or \name{BINARY(size\-In\-Bits)}, then in layers \lite{} and \standard{} they are defined as equivalent to data\-types without size. But in layer \advanced{}, the size parameter must be declared by using property \name{owl:onDatatype} and data\-type facets \name{xsd:maxLength} or \name{xsd:length} \cite{w3c:owl2-syntax,w3c:xmlschema11-2}, for instance:

% \begin{lstlisting}[caption={Size constraint of simple datatypes},label=lst:ifcOWL-simple-types-with-size]
\begin{lstlisting}
# only for Lite, Standard layers
expr:STRING22 owl:sameAs expr:STRING .

# only for Advanced layer
expr:STRING22 owl:equivalentClass
    [   a rdfs:Datatype ;
        owl:onDatatype  xsd:string ;
        owl:withRestrictions ( [ xsd:length "22"^^xsd:integer ] ) ] .
\end{lstlisting}
\end{ontologyRule}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Enumeration Datatype Declarations
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Enumeration Datatype Declarations}
\label{subsec:ifcOWL-enum-types}

An enumeration data\-type represents a set of possible fixed values.

\begin{ontologyRule}[Enumeration data\-types]
Each enumeration data\-type in \textbf{ifc\-OWL} is defined as an \name{owl:Class} which is a subclass of \name{expr:\-Enum\-Class}. All possible values of enumeration data\-types are defined as named individuals. Enumeration data\-type memberships are defined by means of \name{rdf:type} in the \lite{} layer, and by means of \name{owl:oneOf} in \standard{} and \advanced{} layers as shown below:
% \begin{lstlisting}[caption={Enumeration data\-types}, label=lst:ifcOWL-enum-types]
\begin{lstlisting}
# only for Lite layer
ifc:IfcAssemblyPlaceEnum a owl:Class ; rdfs:subclassOf expr:EnumClass .
ifc:SITE, ifc:FACTORY, ifc:NOTDEFINED rdf:type ifc:IfcAddressTypeEnum .

# only for Standard, Advanced layers
ifc:IfcAssemblyPlaceEnum a owl:Class ;  rdfs:subclassOf expr:EnumClass ;
    owl:oneOf ( ifc:SITE ifc:FACTORY ifc:NOTDEFINED ) .
\end{lstlisting}
\end{ontologyRule}

The reason of two different approaches is that enumeration with more than one individual or literal are not supported in OWL 2 EL or QL. In both cases enumeration values are defined as individuals of one class so in terms of reasnoning logic they are equivalent. The enumeration values are defined as individuals because OWL 2 RL supports \name{Object\-One\-Of} to define enumeration of individuals, but not \name{Data\-One\-Of} \cite{w3c:owl2-profiles}. The fact that some values such as \name{ifc:NOT\-DE\-FI\-NED} belong to different enumeration data\-types should not cause any problems since all enumeration values in ifc\-RDF datasets are defined together with a concrete enumeration data\-type (see \ref{subsec:ifcRDF-enum-values}). It is also very easy to get all possible values of an enumeration data\-type by using simple SPARQL queries.

As mentioned above, simple data\-types \name{BOOLEAN} and \name{LOGICAL} are considered by us as enumeration data\-types. But these types and their values must be defined in the EXPRESS namespace, e.g. \name{expr:LOGICAL}, \name{expr:TRUE}, \name{expr:FALSE} or \name{expr:UNKNOWN}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Select Datatype Declarations
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Select Datatype Declarations}
\label{subsec:ifcOWL-select-types}

A select data\-type represents an union of other data\-types. Union members of a select data\-type usually are either entity data\-types, or defined data\-types all together. But there are a few exceptions, for instance, type \name{Ifc\-Trim\-ming\-Se\-lect} is an union of entity data\-type \name{Ifc\-Car\-te\-sian\-Point} and defined data\-type \name{Ifc\-Pa\-ra\-me\-ter\-Va\-lue}.

\begin{ontologyRule}[Select data\-types]
A select data\-type must be converted to an \name{owl:Class} which is a subclass of \name{expr:\-Select\-Class}. Union-member data\-types are defined by means of \name{rdfs:\-subclassOf} in the \lite{} layer, and by means of \name{owl:unionOf} in \standard{} and \advanced{} layers as shown in the listing below.
\end{ontologyRule}

%\begin{lstlisting}[caption={Select data\-types}, label=lst:ifcOWL-select-types]
\begin{lstlisting}
# only for Lite layer
ifc:IfcTrimmingSelect a owl:Class ; rdfs:subclassOf expr:SelectClass .
ifc:IfcCartesianPoint, ifc:IfcParameterValue
    rdfs:subclassOf ifc:IfcTrimmingSelect .

# only for Standard, Advanced layers
ifc:IfcTrimmingSelect a owl:Class ; rdfs:subclassOf expr:SelectClass ;
    owl:unionOf ( ifc:IfcCartesianPoint ifc:IfcParameterValue ) ] .
\end{lstlisting}

Similarly to the above case, two different approaches are used because property \name{owl:unionOf} supported only in OWL 2 RL, but not in OWL 2 EL and OWL 2 QL \cite{w3c:owl2-profiles}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Defined Datatype Declarations
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A defined (declared, named) data\-type is usually specified as equivalent to a simple data\-type or another defined data\-type, with or without a additional constraint. For instance, type \name{Ifc\-Length\-Mea\-sure} is defined as equal to \name{REAL}, and type \name{Ifc\-Po\-si\-ti\-ve\-Length\-Mea\-su\-re} is a set of positive \name{Ifc\-Length\-Mea\-sure} values.

\begin{ontologyRule}[Defined data\-types]
If the defined data\-type is based on a simple data\-type then in ifcOWL it is declared as a \name{owl:Class} which is a subclass of \name{expr:\-Defined\-Class} has a single property \name{rdf:value} with that simple data\-type. If the defined data\-type is based on another defined data\-type then it is declared as a subclass of the \name{owl:Class} of that defined data\-type, for instance:

%\begin{lstlisting}[caption={Defined data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}[aboveskip=3pt]
ifc:IfcLengthMeasure a owl:Class ; rdfs:subClassOf expr:DefinedClass ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty rdf:value;
                      owl:allValuesFrom xsd:double ] .

ifc:IfcPositiveLengthMeasure a owl:Class ;
    rdfs:subClassOf ifc:IfcLengthMeasure .
\end{lstlisting}
\end{ontologyRule}

Defined data\-types are not declared as a \name{rdfs:Datatype}, but \name{owl:Class}es which wrap \name{rdfs:Datatype} values because in contract to simple data types they can be used for declaration of select data\-types (see \ref{subsec:ifcOWL-select-types}).

Defined data\-type \name{Ifc\-Time\-Stamp} is declared as equivalent to \name{INTEGER} but represents so-called Unix time value that can be easily converted to \name{xsd:date\-Time}. So it wraps a \name{xsd:date\-Time} value instead of \name{xsd:integer}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Entity Datatype Declarations
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Entity Datatype Declarations}
\label{subsec:ifcOWL-entity-types}

Entity data\-types are identical to classes in object-oriented programming languages. Each type may have not more than one super\-type that it inherits attributes from. An attribute is defined with a data\-type, a minimum and a maximum cardinalities, can be marked as optional.

\begin{ontologyRule}[Entity data\-types]An entity data\-type is converted to an \name{owl:\-Class} which is subclass of \name{expr:\-Entity\-Class} or its superclass. Entity attributes are defined by means of property \name{owl:on\-Pro\-per\-ty}. If an attribute is optional then its minimum qualified cardinality is 0, otherwise -- 1. Maximum qualified of all attributes are always 1. In case if an attribute is a list or a set of some data\-type then its type is considered as an external aggregated data\-type, for example:
%\begin{lstlisting}[caption={Entity data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}
ifc:IfcProperty a owl:Class ; rdfs:subClassOf expr:EntityClass .

ifc:IfcComplexProperty a owl:Class ; rdfs:subClassOf ifc:IfcProperty ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty ifc:coordinates;
                      owl:allValuesFrom LIST_1_3_IfcLengthMeasure ] ;
    # only for Advanced layer
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty ifc:coordinates;
                      owl:cardinality "1"^^xsd:integer ] .

# only for Standard & Advanced layers
[] a owl:AllDisjointClasses ;
    owl:members ( ifc:IfcComplexProperty ifc:IfcSimpleProperty ) .
\end{lstlisting}
\end{ontologyRule}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Aggregated Datatype Declarations
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Aggregated Type Declarations}
\label{subsec:ifcOWL-aggregated-types}
